---
title: "DATA622_HW4"
author: "Group 5"
date: "`r Sys.Date()`" # Due 5/7/2021
output: 
 html_document:
    toc: true
    toc_float: true
    code_folding: hide
    theme: flatly
    highlight: tango
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
#libraries
library(dplyr)
library(tidyr)
library(ggplot2)
library(inspectdf)
library(corrplot)
library(stats)
library(kableExtra)
library(factoextra)
```

## Authorship

**Group 5:** 

* Don (Geeth) Padmaperuma,
* Subhalaxmi Rout, 
* Isabel R., and
* Magnus Skonberg


# Background

The purpose of this assignment was to explore Clustering, Principal Component Analysis, and Support Vector Machines.

*Clustering* is used as a means of grouping a set of observations in a way that those in the same group are more similar to one another than those in other groups. *Principal Component Analysis* is used to a means of variable reduction where it's our aim to preserve as much useful information as possible. In short, we trade a little accuracy for simplicity. *Support Vector Machines*, on another note, are a supervised learning technique used for classification and regression analysis. We're going to apply it as a means of classification based on one of the variables in our set (suicide).

## Our Approach

We'll perform a relatively in-depth exploratory data analysis (EDA) to better familiarize ourselves with the data at hand and then we'll explore each of the aforementioned methods.

Grouping our patients via k-means / hierarchical clustering and profiling these groups based on shared characteristics, focusing upon a subset of our data (ie. ADHD or mood disorder) in applying Principal Component Analysis, and applying a Support Vector Machine model based on whether or not a patient has attempted suicide.


................................................................................

# The Data

The data at hand is based on a real life research project and deals with a number of measures of mental health and well-being - attention deficit hyperactivity disorder (ADHD), mood / mental disorder (MD), substance abuse, etc. 

Being that our exploration and application of methods can take on many forms and head in a nearly infinite number of directions, the room for creativity is without bounds.

We start by loading in our data and utilizing built-in head() and dim() functions for an early read.
```{r}
#Load in data
data <- read.csv("https://raw.githubusercontent.com/SubhalaxmiRout002/Data-622-Group-5/main/HW%204/ADHD_data.csv")
#head(data)
#dim(data)
```

## Exploratory Data Analysis

From our initial read of our dataset's dimensions and first 6 entries, we observe that we're dealing with a 175 observation x 54 variable dataset. Our initial dataset is broad and may be difficult to manage / make sense of. With the motivation of clarification in mind, we thus break our larger dataset into subsets:

1. **genl**: all identifier variables (race, sex, education), total score variables, and history-related variables.
2. **adhd**: all adhd related question and total score variables.
3. **md**: all mental disorder related question and total score variables.
4. **sub**: all substance abuse related question and total score variables.
5. **hstry**: all violence, abuse, and substance history related variables.

We may end up merging these subsets, pulling individual variables, or exploring individual subsets later but this provides a means of effectively recognizing where related variables lie while simultaneously providing smaller, more manageable dataframes to explore and work upon.

Once these dataframes are created, we see that **adhd** and **md** have total score variables whereas **sub** does not, even though its data (across variables) maintains a consistent scoring system. Thus, we take it upon ourselves to add this variable for sake of consistency:

```{r}
genl <- data[c(1:4,23,39)]
#head(genl) #drop initial, add sub and hstry (all vars or "total score")
#we may want to use these separately (ie. for Q2)
adhd <- data[c(5:23)]
md <- data[c(24:39)]
sub <- data[c(40:45)]
hstry <- data[c(46:54)]
sub$SUB.Total <- rowSums(sub) #add Total column to sub
#Finalize genl dataframe (for EDA)
genl$SUB.Total <- sub$SUB.Total #add SUB.Total to genl
genl <- genl[c(-1)] #drop Initial
genl <- cbind(genl, hstry) #add history variables back in with general
head(genl)
```

We subset our dataframes (as noted above), add a "total" column for **sub**, and finalize our **genl** dataframe by merging the total column for **sub**, dropping the `Initial` variable, and adding all **hstry** variables back in.

We lean on the data dictionary (provided with our .xlsx file) to outline the variables at hand:

* `Age`: quantitative variable representative of the individual's age.
* `Sex`: categorical variable representative of the individual's sex (male-1, female-2).
* `Race`: categorical variable representative of the individual's race (white-1, african american-2, hispanic-3, asian-4, native american-5, other / missing data-6).
* `ADHD.Total`: quantitative variable representative of the individual's total self-report score for ADHD scale. It's the cumulative score for 18 questions with a scoring metric of never-0, rarely-1, sometimes-2, often-3, and very often-4.
* `MD.TOTAL`: quantitative variable representative of the individual's total self-report score for mental disorder questions. It's the cumulative score for 15 questions with a scoring metric of no-0, yes-1; question 3: no problem-0, minor-1, moderate-2, and serious-3.
* `SUB.Total`: quantitative variable representative of the individual's total self-report score for substance abuse related questions. It's the cumulative score across 6 categories with a scoring metric of no use-0, use-1, abuse-2, and dependence-3.
* `Court.order`: categorical variable representative of whether the individual's case was court ordered (No-0, Yes-1).
* `Education`: categorical variable representative of the individual's level of education (1-12 grade, 13+ college).
* `Hx.of.Violence`: categorical variable representative of whether the individual has a history of violence (No-0, Yes-1).
* `Disorderly.Conduct`: categorical variable representative of whether the individual has a record of disorderly conduct (No-0, Yes-1).
* `Suicide`: categorical variable representative of whether the individual has attempted suicide in the past (No-0, Yes-1).
* `Abuse`: categorical variable representative of whether the individual has a history of abuse (No-0, Physical (P)-1, Sexual (S)-2, Emotional (E)-3, P&S-4, P&E-5, S&E-6, P&S&E-7).
* `Non.subst.Dx`: categorical variable representative of whether the individual has a non-substance diagonosis (0 – none; 1 – one; 2 – More than one).
* `Subst.Dx`: categorical variable representative of whether the individual has a substance diagonosis (0 – none; 1 – one Substance-related; 2 – two; 3 – three or more).
* `Psych.meds.`: categorical variable representative of whether the individual has been prescribed psychiatric medication (0 – none; 1 – one psychotropic med; 2 – more than one psychotropic med).

The resulting dataframe (and variables) are shown above and at this point we're prepared to more deeply explore our **genl** dataset.

To do so, we utilize the built-in summary() and glimpse() functions:

```{r}
summary(genl)
glimpse(genl)
```

From the above outputs we can extend that:

* the average `Age` of respondents is ~40, there are slightly more male than female respondents, and the average education level ~12 (finishing high school).
* the average `ADHD,Total` was ~34, the average `MD.TOTAL` was ~10, and the averge `SUB.Total` was ~4.
* `Disorderly.Conduct` appears to be more common than `Hx.of.Violence` which in turn is more common than `Court.order`.
* approximately 1/3 of respondents have attempted `Suicide` and some form of abuse appears to be common (ie. Physical).
* the average `Non.subst.Dx`  was ~0.4 while the average `Subst.Dx` was ~1.1 giving the impression that we're dealing with more substance diagnoses that non-substance diagnoses.
* `Psych.meds` is missing the majority of its values and NA values will have to be accounted for (in general).

Based on this information, we drop `Psych.meds`, drop remaining NA values, and observe the affect on our dataset:

```{r}
genl <- genl[c(-15)] #drop Psych.meds
genl <- genl %>% tibble() %>% drop_na() #drop NA values
#revisit genl
dim(genl) #175 - 142 = 33 dropped rows
colSums(is.na(genl)) #verify no NAs
```

Our dataset has been reduced from 175 observations x 15 variables to 142 observations x 14 variables and from numerous NA values across multiple variables to no NA values.

With NA values dealt with, we ensure each variable's of the proper type and then visualize our variable distributions based on whether they're categorical or numeric: 

```{r}
#convert features to factor
genl$Sex <- as.factor(genl$Sex)
genl$Race <- as.factor(genl$Race)
genl$Court.order <- as.factor(genl$Court.order)
genl$Education <- as.factor(genl$Education)
genl$Hx.of.Violence <- as.factor(genl$Hx.of.Violence)
genl$Disorderly.Conduct <- as.factor(genl$Disorderly.Conduct)
genl$Suicide <- as.factor(genl$Suicide)
genl$Abuse <- as.factor(genl$Abuse)
genl$Non.subst.Dx <- as.factor(genl$Non.subst.Dx)
genl$Subst.Dx <- as.factor(genl$Subst.Dx)
#conver features to numeric
genl$Age <- as.numeric(genl$Age)
genl$ADHD.Total <- as.numeric(genl$ADHD.Total)
genl$MD.TOTAL <- as.numeric(genl$MD.TOTAL)
genl$SUB.Total <- as.numeric(genl$SUB.Total)
#head(genl) #verify conversions
#visualize categorical distributions as a table
fig1 <- inspectdf::inspect_imb(genl)
fig1
```

From the categorical variable table above we see that ~92% of respondent's cases were not court ordered, ~74% of respondents do not have a history of violence, ~70% of respondents have a history of disorderly conduct, ~68% of respondents have not attempted suicide, ~65% of respondents do not have a non-substance diagnosis, ~63% of respondents do not have a history of abuse, ~56% of respondents are African American, ~54% of respondents are male, ~39% of respondents have completed *at least* high school, and ~39% of respondents have one substance related diagnosis. *The fact that `Education` and `Subst.Dx` have the same pcnt values is interesting and worth noting.*

Moving on to our numeric distributions:
```{r}
#visualize numeric distributions 
fig2 <- inspectdf::inspect_num(genl) %>% 
    show_plot()
fig2
```

From the histograms above we see that:

* `ADHD.Total` has a left skewed normal distribution with a peak ~45.
* `Age` appears to be bimodal with peaks ~25 and ~50.
* `MD.TOTAL` has a non-normal distribution with a peak concentrated > 15 (at the far right).
* `SUB.Total` appears to have a multi-modal distribution with its largest peak at ~3.

With insight gained via variable distributions, we move on to observing the pair plot and correlation matrix for numeric variables:

```{r}
#paired plot of numeric variables
pairs(genl %>% select_if(is.numeric))
#Correlation matrix for numeric variables
genl_corr <- genl %>%
    select_if(is.numeric) %>%
    cor()
corrplot(genl_corr, title="Numeric Variable Correlation",type = "lower", mar=c(0,0,1,0))
```

The pair plot and correlation matrix highlight the fact that `ADHD.Total` and `MD.TOTAL` appear to have strong correlation. We'll note this fact, while carrying both variables forward.


................................................................................

# 1. Clustering

We elected to utilize **hierarchical clustering** being that it's more adaptive than k-means clustering. For k-means clustering, we have to pre-define our k. Whereas for hierarchical clustering both the value of k and the location of our centroids (mid-point between all data points in the group) is determined as a part of the process.

We proceed by computing the distances between all data points (using Euclidean distance), clustering based on `distances` based on centroid distance *and* the variance in each cluster, and then plotting the corresponding dendrogram:

```{r}
set.seed(123)
#compute distances between all data points (using euclidean distance)
distances <- dist(genl, method = 'euclidean')
#cluster based on distances using centroid distance as well as variance in clusters
clusterHealth <- hclust(distances, method = 'ward.D') #ward.D accounts for centroid distance + variance
#plot the dendrogram
plot(clusterHealth)
```

Based on our dendrogram, it appears that the ideal number of clusters is either 3, 4, or 5. **We proceed with 5 clusters** being that it's the higher of the three and we'd like to better represent the variation we observed during our exploratory data analysis. We want to better capture the distinct characteristics of our clustered groups.

We "cut the tree" and demarcate our groups based on the desired number of clusters, calculate each numeric variables' mean value for each cluster, and then output the resulting table:
```{r}
clusterGroups <- cutree(clusterHealth, 5)
#clusterGroups #verify the breakdown per observation
#Mean value for numeric variables
#tapply(genl$Age, clusterGroups, mean, simplify=TRUE) #avg age for each cluster
#tapply(genl$ADHD.Total, clusterGroups, mean) #avg ADHD score for each cluster
#tapply(genl$MD.TOTAL, clusterGroups, mean) #avg MD score for each cluster
#tapply(genl$SUB.Total, clusterGroups, mean) #avg SUB score for each cluster
#Extract corresponding values (to place in easy to understand form):
avg_age <- c(27.06, 48.95, 41.42, 47.4, 25.22)
avg_adhd <- c(49.34, 53.52, 34.12, 15.93, 6.56)
avg_md <- c(12.84, 12.71, 10.58, 7.27, 4.11)
avg_sub <- c(4.13, 3.28, 4.1, 4.17, 3.44)
#Build kable table
group_names <- c("Group 1", "Group 2", "Group 3", "Group 4", "Group 5")
df <- data.frame(group_names, avg_age, avg_adhd, avg_md, avg_sub)
df %>% 
    kbl() %>%
    kable_styling(latex_options = "striped")
```

Based on the output above we'd profile our groups as follows:

* Group 1: **Young and Distressed**. This is the 2nd youngest group (late 20s) with the 2nd highest incidence of ADHD and the highest incidence of mental disorder.
* Group 2: **Old and Unfocused**. This is the oldest group (late 40s) with the highest incidence of ADHD and the 2nd highest incidence of mental disorder.
* Group 3: **Mid Age and Mid Tiered**. This group is "middle of the run". They're toward their middle years (early 40s), and have moderate incidence of ADHD and mental disorder.
* Group 4: **Old and Stable**. This is the 2nd most stable group. They're in their  late 40s and have low incidence of ADHD and mental disorder.
* Group 5: **Young and Driven**. To contrast the **Young and Distressed** group, this group is young and stable. This is the most stable group. They're in their mid 20s and have the lowest incidence of ADHD and mental disorder.

It was interesting to profile our groups and observe that there appeared to be 2 tiers in the earlier and later years. It was also interesting to see that all groups had a relatively low incidence of substance use / dependency. Due to its apparent low impact as a differentiating factor for our clusters it was not discussed.


................................................................................

# 2. Principal Component Analysis

Here we are using built-in R function *prccomp()* to perform the *Principal Component Analysis*. 

PCA can be applied only on numerical data. We have few Categorical variables in this dataset that need to be converted to numberical category. 

```{r}
# Check the data class
sapply(genl, class)
```
As we can see there are only 4 numeric data types and rest are factors. We need to convert all the factors into numeric.

```{r}
cols.num <- c("Sex","Race","Court.order","Education","Hx.of.Violence", "Disorderly.Conduct", "Suicide","Abuse","Non.subst.Dx","Subst.Dx")
genl[cols.num] <- sapply(genl[cols.num],as.numeric)
```


```{r}
# Check to see if all our variables in numeric.
sapply(genl, class)
```

```{r}
# Compute principal component analysis.
result.pca <- prcomp(genl, scale = TRUE)
```

Show the percentage of variances explained by each principal component.

```{r}
# Visualize eigenvalues (scree plot). 
fviz_eig(result.pca)
```
Individuals with a similar profile are grouped together.

```{r}
# Graph of individuals
fviz_pca_ind(result.pca,
             col.ind = "cos2", # Color by the quality of representation
             gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"),
             repel = TRUE     # Avoid text overlapping
             )
```

Positive correlated variables point to the same side of the plot. Negative correlated variables point to opposite sides of the graph.

```{r}
# Graph of variables
fviz_pca_var(result.pca,
             col.var = "contrib", # Color by contributions to the PC
             gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"),
             repel = TRUE     # Avoid text overlapping
             )
```
```{r}
# Biplot of individuals and variables
fviz_pca_biplot(result.pca, repel = TRUE,
                col.var = "#2E9FDF", # Variables color
                col.ind = "#696969"  # Individuals color
                )
```

### Access to the PCA results
```{r}
# Eigenvalues
eig.val <- get_eigenvalue(result.pca)
eig.val
  
# Results for Variables
res.var <- get_pca_var(result.pca)
res.var$coord          # Coordinates
res.var$contrib        # Contributions to the PCs
res.var$cos2           # Quality of representation 
# Results for individuals
res.ind <- get_pca_ind(result.pca)
res.ind$coord          # Coordinates
res.ind$contrib        # Contributions to the PCs
res.ind$cos2           # Quality of representation 
```

................................................................................

# References

In completing this assignment, reference was made to the following:

* MIT OpenCourseWare. (2018). **An Introduction to Clustering - Video 7: Hierarchical Clustering in R** [video]. Retrieved from: https://www.youtube.com/watch?v=GPOUGpF-Sno [hierarchical clustering]
© 2021 GitHub, Inc.

* Statistical tools for high-throughput data analysis. (2017)
**Principal Component Analysis in R: prcomp vs princomp** Retrieved from: http://www.sthda.com/english/articles/31-principal-component-methods-in-r-practical-guide/118-principal-component-analysis-in-r-prcomp-vs-princomp/