---
title: "DATA622_HW3"
author: "Group 5 - Don (Geeth) Padmaperuma,Subhalaxmi Rout, Isabel R., Magnus Skonberg"
date: "4/9/2021"
output: 
 html_document:
    toc: true
    toc_float: true
    code_folding: hide
    theme: flatly
    highlight: tango
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r library, comment=FALSE, warning=FALSE, message=FALSE }
library(tidyverse)
library(caret)
library(palmerpenguins)
library(e1071) 
library(caTools)
library(ggplot2)
library(GGally)
library(ggplot2) 
library(MASS) 
library(mvtnorm)
```

# K-nearest neighbor (KNN)
1. Please use K-nearest neighbor (KNN) algorithm to predict the species variable. Please
be sure to walk through the steps you took. (40 points)

# Decision Trees
2. Please use the attached dataset on loan approval status to predict loan approval using
Decision Trees. Please be sure to conduct a thorough exploratory analysis to start the
task and walk us through your reasoning behind all the steps you are taking. (40 points)

# Random Forests
3. Using the same dataset on Loan Approval Status, please use Random Forests to predict
on loan approval status. Again, please be sure to walk us through the steps you took to
get to your final model. (50 points)

# Gradient Boosting
4. Using the Loan Approval Status data, please use Gradient Boosting to predict on the
loan approval status. Please use whatever boosting approach you deem appropriate;
but please be sure to walk us through your steps. (50 points)

# Model Comparison (Decision Trees vs. Random Forests vs. Gradient Boosting)
5. Model performance: please compare the models you settled on for problem # 2 – 4. Comment on their relative performance. Which one would you prefer the most? Why? (20 points)
